---
title: "Homework 7"
author: "Andrew Boschee"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F,warning=F,echo=F,fig_height=10,fig_width=7,cache = F)
```
*No collaborators, *
*Outside Resources: R Graphics Cookbook, Rdocumentation.com* 

```{r, echo=FALSE}
#install.packages('survival')
library(survival)
library(ggplot2)
#install.packages('survminer')
library(survminer)
library(HSAUR3)
#install.packages('ISwR')
library(ISwR)
#install.packages('coin')
library(coin)
library(dplyr)
library(knitr)
library(gridExtra)
library(partykit)
```

1. An investigator collected data on survival of patients with lung cancer at Mayo Clinic. The investigator would like you, the statistician, to answer the following questions and provide some graphs. Use the \textbf{cancer} data located in the \textbf{survival} package.

a. What is the probability that someone will survive past 300 days?
See Table 1 below...

```{r, echo=FALSE, fig.width=10, fig.height=5}
# load dataset
data(cancer)

# build first model of survival
survModel <- survfit(Surv(time, status == 2)~1, data = cancer)

# pull survival probability from summary
survProb300 <- summary(survModel, time = 300)[6]
```

b. Provide a graph, including 95% confidence limits, of the Kaplan-Meier estimate of the entire study.


```{r, echo=FALSE, fig.width=10, fig.height=5}

# base r plot with confidence limits
plot(survModel, xlab = 'Time (Days)', ylab = 'Survival Probability', main = 'Probability of Survival by # of Days',
     col = c('black','red','red'))
legend('topright', c('Survival by # of Days', 'CI Limits'), col = c('black','red','red'), lty = c(1,2,2))
```

```{r, echo=FALSE, fig.width=10, fig.height=5}

# replicate of prior plot with ggsurvplot
ggsurvplot(survModel, data = cancer, conf.int = TRUE, palette = c('black'),
           title = 'Probability of Survival by # of Days')
```

```{r,echo=FALSE, fig.width=10, fig.height=5}
# print probability of survival
kable(survProb300, col.name = 'Survival Probability', caption = 'Probability of Survival Past 300 Days')

```   
    
c. Is there a difference in the survival rates between males and females? Provide a formal statistical test with a p-value and visual evidence.

There is a significant difference when it comes to gender and survival rates. With 67% survival for females and 44% survival for males and a small p-value in Table 3, gender plays a major role in lung cancer survival. 

Saw this as a good time to look into the Cox Proportional Hazards model and see what else plays key roles outside of just age and gender that we are exploring. This brought up the ph.ecog variable that is the most signficant along with ph.karno. A conditional inference tree is also provided to show visualization of gender along with the most signficant variable at the first split of the tree.
    
```{r, echo = FALSE}
# make subsets by gender
survMale <- subset(cancer, cancer$sex ==1)
survFemale <- subset(cancer, cancer$sex ==2)

# turn gender to factor level
survMale$sex<- as.factor('male')
survFemale$sex <- as.factor('female')

# build survival model for each gender
survMaleModel <- survfit(formula = Surv(time, status ==2) ~ 1, data = survMale)
survFemaleModel <- survfit(formula = Surv(time, status == 2) ~ 1, data = survFemale)
```

```{r, echo = FALSE, fig.width=10, fig.height=5}
# store probability in variable for kable output
maleSurvRate <- summary(survMaleModel, time = 300)[6]
femaleSurvRate <- summary(survFemaleModel, time = 300)[6]

# combine reults to table for output formatting
survDF <- cbind(maleSurvRate, femaleSurvRate)

kable(survDF, row.names = FALSE, col.names = c('Male', 'Female'), caption = 'Probability 300+ Days of Survival by Gender')
```

```{r, echo = FALSE, fig.width=10, fig.height=5}
# turn gender to factor
cancer$sex <- as.factor(cancer$sex)

# get p value from lgorank test and output
cancerPVal <- pvalue(logrank_test(Surv(time, status == 2) ~ sex, data = cancer))

kable(cancerPVal, col.names = 'P-Value', caption = 'P-Value of Survival Significance by Gender')


```
    
```{r, echo = FALSE, fig.width=10, fig.height=5}

# base R plot of survival probability by gender
layout(matrix(c(1:2), ncol = 2))
plot(survMaleModel, xlab = 'Time (Days)', ylab = 'Survival Probability', main = 'Probability of Survival by # of Days - Male',
     col = c('black','red','red'))
legend('topright', c('Survival by # of Days', 'CI Limits'), col = c('black','red','red'), lty = c(1,2,2))

plot(survFemaleModel, xlab = 'Time (Days)', ylab = 'Survival Probability', main = 'Probability of Survival by # of Days - Female',
     col = c('black','red','red'))
legend('topright', c('Survival by # of Days', 'CI Limits'), col = c('black','red','red'), lty = c(1,2,2))
```
    
```{r, echo = FALSE, fig.width=10, fig.height=5}

#replicate prior plot with ggsurvplot
ggsurvplot(survMaleModel, data = survMale, conf.int = TRUE, palette = c('black'),
           title = 'Probability of Survival by # of Days - Male')


```

```{r, echo = FALSE, fig.width=10, fig.height=5}
# replicate prior plot with ggsurvplot
ggsurvplot(survFemaleModel, data = survFemale, conf.int = TRUE, palette = c('black'),
           title = 'Probability of Survival by # of Days - Female')
```

\pagebreak

## Cox Proportional Hazard Regression

```{r, echo = FALSE}
# get surmmary of cox hazard regression
coxHazard <- coxph(Surv(time, status ==2) ~., data = cancer)
summary(coxHazard)

```
```{r, echo = FALSE, fig.width=10, fig.height=5}
# plot conditional inference tree
decTree <- ctree(Surv(time, status ==2) ~., data = cancer)
plot(decTree)

```

d. Is there a difference in the survival rates for the older half of the group versus the younger half? Provide a formal statistical test with a p-value and visual evidence.

After finding the median age, the data was split into to subsets, 'older' and 'younger', with survfit() applied to each. Surprisingly, there is not a significant impact on survival rate according to the model with a p-value at .17. With the median age at 63, I was a little on the edge of whether that is a high enough age where the study looking at survival of 300 days would be impacted signifcantally at that age.
    
```{r, echo = FALSE}

#find median age
medianAge <- median(cancer$age)

# split data between above and below median age
cancer$ageLevel <- ifelse(cancer$age > medianAge, 'older','younger')
cancer$ageLevel <- as.factor(cancer$ageLevel)


olderDF <- subset(cancer, cancer$ageLevel == 'older')
youngerDF <- subset(cancer, cancer$ageLevel =='younger')

# build model on both age groups
olderSurvModel <- survfit(formula = Surv(time, status ==2) ~ 1, data = olderDF)
youngerSurvModel <- survfit(formula = Surv(time, status == 2) ~ 1, data = youngerDF)

```
```{r, echo = FALSE, fig.width=10, fig.height=5}

# get summaries and put into table for kable output comparison
olderSurvRate <- summary(olderSurvModel, time = 300)[6]
youngerSurvRate <- summary(youngerSurvModel, time = 300)[6]

survDF2 <- cbind(olderSurvRate, youngerSurvRate)

kable(survDF2, row.names = FALSE, col.names = c('Older', 'Younger'), caption = 'Probability 300+ Days of Survival by Age Group')
```

```{r, echo = FALSE, fig.width=10, fig.height=5}

# calc pval with logrank test for agelevel
cancerPVal2 <- pvalue(logrank_test(Surv(time, status == 2) ~ ageLevel, data = cancer))

kable(cancerPVal2, col.names = 'P-Value', caption = 'P-Value of Survival Significance by Age Group')
```


```{r, echo = FALSE, fig.width=10, fig.height=5}

# base R plot of age differnece survival 
layout(matrix(c(1:2), ncol = 2))
plot(youngerSurvModel, xlab = 'Time (Days)', ylab = 'Survival Probability', main = 'Probability of Survival by # of Days - Younger',
     col = c('black','red','red'))
legend('topright', c('Survival by # of Days', 'CI Limits'), col = c('black','red','red'), lty = c(1,2,2))

plot(olderSurvModel, xlab = 'Time (Days)', ylab = 'Survival Probability', main = 'Probability of Survival by # of Days - Older',
     col = c('black','red','red'))
legend('topright', c('Survival by # of Days', 'CI Limits'), col = c('black','red','red'), lty = c(1,2,2))
```

```{r, echo = FALSE, fig.width=10, fig.height=5}

# replicate base R pltos
ggsurvplot(youngerSurvModel, data = youngerDF, conf.int = TRUE, palette = c('black'),
           title = 'Probability of Survival by # of Days - Younger')
```

```{r, echo = FALSE, fig.width=10, fig.height=5}

# replicate Base R plot
ggsurvplot(olderSurvModel, data = olderDF, conf.int = TRUE, palette = c('black'),
           title = 'Probability of Survival by # of Days - Older')
```

2. A healthcare group has asked you to analyse the \textbf{mastectomy} data from the \textbf{HSAUR3} package, which is the survival times (in months) after a mastectomy of women with breast cancer. The cancers are classified as having metastasized or not based on a histochemical marker. The healthcare group requests that your report should not be longer than one page, and must only consist of one plot, one table, and one paragraph. Do the following:

a. Plot the survivor functions of each group only using GGPlot, estimated using the Kaplan-Meier estimate.
    
```{r, echo = FALSE, fig.width=10, fig.height=5}
# load data
data(mastectomy)

# make event binary
mastectomy$event <- ifelse(mastectomy$event == TRUE, 2, 1)

# build surv movel on data
mastectomySurv <- survfit(formula = Surv(time, event == 2) ~ metastasized, data = mastectomy)
#head(mastectomy)

```

```{r, echo = FALSE, fig.width=10, fig.height = 5}
# plot output
ggsurvplot(mastectomySurv, data = mastectomy, conf.int = TRUE,
           title = 'Mastectomy Survival')
```

b. Use a log-rank test to compare the survival experience of each group more formally. Only present a formal table of your results. 

```{r, echo = FALSE}
# build lograndtuest and output sumamry
logrank_test(Surv(time, event == 2)~ metastasized, data = mastectomy)

```

    
c. Write one paragraph summarizing your findings and conclusions. 

The P-Value is not significant at the level of .05 and our null hypothesis can be rejected. Looking at the graph the survival probability is higher when not having metastasized in comparison to having metastasized. However, when looking at the plot, we can see that the range at a 95% confidence interval is very large not having metastasized in comparison to the range of the having metastasized. If we were testing the hypothesis at a different level, .1 for example, our final determination would be that there is not enough evidence to reject the null hypothesis with a p-value of .06194. With this model being applied to a health situation, especially regarding survival, I believe further studies be performed to explore this topic.
